---
title: "The {ttx} Package"
author: "Paul Murrell"
output: 
  rmarkdown::html_vignette:
    number_sections: true
bibliography: ttx-intro.bib
vignette: >
  %\VignetteIndexEntry{The {ttx} Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = ''
)
```

*The {ttx} package is a front end to the 
[`ttx` utility](https://fonttools.readthedocs.io/en/latest/ttx.html).
It provides functions for extracting font tables from TrueType
and OpenType fonts, plus functions for extracting some specific font and glyph
metric information from the font tables.*

# Font tables

A font is essentially a collection of *glyphs* or shapes, some of which
correspond to letters, but some of which correspond to 
symbols such as punctuation or
accents, etc.
A font file contains lots of information about glyphs, including:
a description of the outline of each glyph;
the *metrics* for each glyph (width, height, descent, etc);
and the correspondence between glyphs and character sets, like UNICODE.

All of this information is stored in a set of tables.
For example, TrueType and OpenType fonts have an 
[`hmtx` table](https://developer.apple.com/fonts/TrueType-Reference-Manual/RM06/Chap6hmtx.html) that contains metric information for the horizontal
typesetting of glyphs.
TrueType and OpenType fonts also have a 
[`cmap` table](https://learn.microsoft.com/en-us/typography/opentype/spec/cmap)
that contains mappings between character codes and glyphs.

# Extracting font tables

The main function of the 
`ttx` utility, which is part of the Python fontTools library
[@fonttools], is to convert binary font files into an XML format.
It can also convert back again.

The {ttx} package uses `ttx` to extract font tables into XML files
and then reads the resulting XML into R, using the
{xml2} package [@pkg-xml2].
For example, the following code calls the `ttxGetTable()` function
to extract the "name" table from a Montserrat font file.  This
contains human-readable names for information about the font.

```{r}
library(ttx)
```

```{r}
fontFile <- system.file("fonts", "Montserrat", "static", 
                        "Montserrat-Medium.ttf",
                        package="grDevices")
nameTable <- ttxGetTable("name", fontFile)
nameTable
```

# Extracting information from tables

The result returned by `ttxGetTable()` 
is an `"xml_document"`.  The following code shows the entire XML structure
of this table.

```{r}
cat(as.character(nameTable), sep="\n")
```

We can extract information from an `"xml_document"` using functions
from the {xml2} package.
For example, 
the following code extracts the text content from
the `nameRecord` element with a `nameID` 
attribute value of `1`.
This corresponds to the font family name.

```{r}
library(xml2)
```

```{r}
familyElement <- xml_find_first(nameTable, "//namerecord[@nameID = '1']")
xml_text(familyElement)
```

# Extracting font information

The {ttx} package provides some convenience functions for extracting
common font information.
For example, the following code extracts the font family, font weight,
and font style for the Montserrat font.

```{r}
family <- ttxFontFamily(fontFile)
family
weight <- ttxFontWeight(fontFile)
weight
style <- ttxFontStyle(fontFile)
style
```

# Extracting glyph information

The {ttx} package also provides some convenience functions
for extracting information about individual glyphs.
For example, the following code extracts the width and left-side bearing
for glyph 1654 (a florin character) from the Montserrat font,
plus a bounding box for the same glyph.

```{r}
glyph <- 1654
width <- ttxGlyphWidth(glyph, fontFile)
width
bbox <- ttxGlyphBounds(glyph, fontFile)
bbox
```

The following code uses that information to draw a diagram
of the glyph metrics.  The grey horizontal line represents the
baseline on which we are laying out text.
The left dot is the origin at which
the florin 
glyph is drawn and the right dot shows the origin where the next glyph
would be drawn in a horizontal layout of text.
The dashed bounding box shows that the rendered glyph actually 
extends backwards past its origin and also forwards past the origin
of the next glyph.

```{r}
library(grid)
```

```{r fig.width=1, fig.height=1, dpi=200}
grid.newpage()
size <- 50
left <- unit(.25, "npc")
baseline <- unit(.25, "npc")
convert <- function(x) {
    size*x/1000
}
w <- convert(width)
bb <- convert(bbox)
grid.segments(left - unit(size/2, "pt"), baseline,
              left + unit(size, "pt"), baseline,
              gp=gpar(col="grey"))
grid.circle(x=unit.c(left, left + unit(w["width"], "pt")),
            y=baseline, r=unit(.5, "mm"), 
            gp=gpar(col="black", fill="black"))
font <- glyphFont(fontFile, 0, family, weight, style)
info <- glyphInfo(id=glyph, x=0, y=0, 
                  font=1, size=size, fontList=glyphFontList(font),
                  width=size, height=size)
grid.glyph(info, left, baseline, hjust=0, vjust=0, gp=gpar(col="grey"))
grid.segments(left, baseline,
              left + unit(rep(c(0, w["width"]), each=2), "pt"),
              baseline)
grid.rect(left + unit(bb["xmin"], "pt"),
          baseline + unit(bb["ymin"], "pt"),
          unit(bb["xmax"] - bb["xmin"], "pt"),
          unit(bb["ymax"] - bb["ymin"], "pt"),
          just=c("left", "bottom"),
          gp=gpar(fill=NA, lty="dotted"))
```

# Caching font tables

The extraction of font tables into XML files, the reading of those
XML files into R as `"xml_document"`s, and the accessing of elements and 
attributes of an `"xml_document"` using {xml2} are all expensive
operations.  

In order to limit this expense, 
the {ttx} package maintains a cache of XML files and automatically converts 
the `"xml_document"`s into data frames, which are also cached as RDS files.
This means that, within a single R session, each table from each font
should only be extracted (and converted to a data frame) at most once.

If the `"ttx.cacheDir"` option is set (and the corresponding
directory exists), the {ttx} package will 
use that to directory to cache XML files and RDS files
so that expensive extractions will not be repeated 
in new R sessions.

# References
